---
title: "Migrating from SELMA API v2 to v3"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Migrating from SELMA API v2 to v3}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

SELMA v3 is a significant API revision. This vignette documents every breaking
change that affects selmaR users, and explains how the package handles them
transparently.

## Quick start

```{r}
# v2 (explicit)
con <- selma_connect(api_version = "v2")

# v3 (explicit)
con <- selma_connect(api_version = "v3")

# Auto-detect (default) — probes the API at connect time
con <- selma_connect()
```

The detected version is stored on the connection object and drives all
subsequent requests:

```{r}
con$api_version  # "v2" or "v3"
print(con)       # shows version in summary
```

---

## Structural changes

| | v2 | v3 |
|---|---|---|
| URL path prefix | `/app/` | `/api/` |
| Pagination member key | `hydra:member` | `member` |
| Pagination total key | `hydra:totalItems` | `totalItems` |
| Auth endpoint | `api/login_check` | `api/auth` |
| Field casing (raw API) | Inconsistent (camelCase / PascalCase) | snake_case |

**All of the above are handled automatically** — you do not change any code for
them. When your SELMA instance upgrades, re-run `selma_connect()` and selmaR
auto-detects the new version and adjusts path prefixes, pagination keys, and
auth endpoints accordingly.

---

## JSON-LD metadata: `@id`, `@type`, `@context`

SELMA uses Hydra JSON-LD, which annotates every API response with three
metadata fields:

| JSON-LD field | Purpose |
|---|---|
| `@id` | IRI of the record (e.g. `"/api/students/42"`) |
| `@type` | JSON-LD type (e.g. `"Student"`) |
| `@context` | JSON-LD context URL |

These are internal API machinery — they carry no useful data for analytics.
selmaR drops all three **before** applying `clean_names()`. This means:

- `id` in every returned tibble is always the **integer primary key** — clean and joinable
- `@type` and `@context` are never visible in output
- You never need to work around JSON-LD artifacts

This applies identically to v2 and v3.

---

## IRI foreign key references (v3 only)

In v3, foreign key columns reference related records as IRI strings rather
than plain integers:

```
# Raw v3 API response — enrolments
{
  "id": 101,
  "student": "/api/students/42",
  "intake":  "/api/intakes/10"
}
```

selmaR strips the IRI path and retains only the trailing ID segment as a
**character string**:

```{r}
enrolments <- selma_enrolments(con)

# id is integer (primary key)
# student, intake are character ("42", "10")
str(enrolments[, c("id", "student", "intake")])
#> $ id      : int  101 102 ...
#> $ student : chr  "42" "99" ...
#> $ intake  : chr  "10" "10" ...
```

This stripping is automatic and version-aware — v2 foreign keys are already
integers and are left untouched.

### Joining v3 data

Because `id` columns are integers and IRI-derived FK columns are characters,
you need a type cast when joining:

```{r}
library(dplyr)

# Option 1: cast the FK to integer before joining
enrolments |>
  mutate(student_id = as.integer(student)) |>
  left_join(students, by = c("student_id" = "id"))

# Option 2: cast the PK to character
enrolments |>
  left_join(
    mutate(students, id = as.character(id)),
    by = c("student" = "id")
  )
```

The convenience join helpers (`selma_join_students()`, `selma_join_intakes()`,
etc.) handle this cast for you.

---

## Stable function names across API versions

Many SELMA v3 endpoints were renamed from v2. selmaR resolves these
transparently via an internal alias table so all function names work regardless
of which API version your instance runs.

Examples of functions that route to different endpoint paths by version:

| selmaR function | v2 endpoint | v3 endpoint |
|---|---|---|
| `selma_ethnicities()` | `ethnicities` | `new_zealand_ethnicities` |
| `selma_nz_iwis()` | `nz_iwis` | `new_zealand_iwis` |
| `selma_nz_residential_statuses()` | `n_z_residential_statuses` | `new_zealand_residential_statuses` |
| `selma_secondary_schools()` | `secondary_schools` | `new_zealand_secondary_schools` |
| `selma_secondary_quals()` | `secondary_quals` | `new_zealand_secondary_school_qualifications` |
| `selma_withdrawal_reason_codes()` | `withdrawal_reason_codes` | `withdrawal_reasons` |
| `selma_enr_status_codes()` | `enr_status_codes` | `enrolment_statuses` |
| `selma_classes()` | `classes` | `sys_classes` |
| `selma_contacts()` | `student_contacts` | `student_contact_associations` |

No code changes are required when your SELMA instance upgrades.

---

## v2-only endpoints

Some SELMA v2 endpoints have no v3 equivalent. Functions wrapping these endpoints
detect the connection version and abort with an informative message rather than
sending a request that would fail silently:

```{r}
# On a v3 connection, these abort immediately with a clear error
selma_student_custom_fields("123", con)
#> Error: selma_student_custom_fields() uses a v2-only endpoint with no v3 equivalent.
#> ℹ Check the v3 API spec for an alternative endpoint.

selma_enrolment_custom_fields("456", con)
selma_component_custom_fields("456", con)
selma_milestones("42", con)
```

### `selma_intake_enrolments()` — v2 only

The `/app/intake_enrolments` endpoint was removed in v3. Use the `filter`
argument on `selma_enrolments()` instead:

```{r}
# v2
ie <- selma_intake_enrolments(intake_id = 123, con = con)

# v3 equivalent
enrolments <- selma_enrolments(con, filter = list(intake = "123"))
```

---

## Server-side filtering — `filter = list()`

All core entity functions (`selma_students`, `selma_enrolments`,
`selma_intakes`, `selma_components`, `selma_programmes`) and all functions
created with `make_entity_fetcher()` accept a `filter` list for server-side
filtering.

Parameter names are validated against the OpenAPI spec for your API version —
unknown parameters emit a warning and are dropped rather than silently sending
a wrong request to the API.

```{r}
# v3 — filter students by first name
smiths <- selma_students(con, filter = list(surname = "Smith"))
alices <- selma_students(con, filter = list(first_name = "Alice"))

# v3 — filter enrolments by intake or student
enrs <- selma_enrolments(con, filter = list(intake = "123"))
enrs <- selma_enrolments(con, filter = list(student = "456"))

# v3 — components for a specific enrolment
comps <- selma_components(con, filter = list(enrolment = "456"))

# v3 — intakes starting after a date
upcoming <- selma_intakes(con, filter = list(`start_date[after]` = "2025-01-01"))
```

To see valid filter parameters for a given function and API version:

```{r}
selmaR:::.selma_schemas$v3$students$params
selmaR:::.selma_schemas$v3$enrolments$params
```

### Parameter names differ between versions

The API parameter names (not the R function names) changed between v2 and v3.
Pass the parameter name appropriate for your API version:

| Entity | v2 parameter | v3 parameter |
|---|---|---|
| students — first name | `forename` | `first_name` |
| students — email | `email1` | `email_primary` |
| students — date of birth | `dob` | `date_of_birth` |
| enrolments — by intake | *(not supported)* | `intake` |
| enrolments — by student | *(not supported)* | `student` |
| components — by enrolment | `enrolid` | `enrolment` |
| intakes — start date range | `intakestartdate[after]` | `start_date[after]` |

---

## Querying endpoints not yet wrapped

SELMA v3 has over 200 collection endpoints. selmaR does not wrap all of them.
Use `make_entity_fetcher()` to create a function for any endpoint, or
`selma_get()` for a one-off query:

```{r}
# Create a reusable function
selma_events     <- make_entity_fetcher("events")
selma_placements <- make_entity_fetcher("placements")
selma_visas      <- make_entity_fetcher("visas")

events <- selma_events(con)

# One-off query
passports <- selma_get(con, "passports")
documents <- selma_get(con, "documents")
```

The endpoint name should be the v3 path segment (e.g. `"events"` for
`/api/events`). The returned function has the same `filter`, `cache`,
`cache_dir`, `cache_hours`, and `items_per_page` arguments as the named
entity functions.

---

## Field name changes

The returned tibble column names differ between versions because v3 changed
field names. selmaR surfaces what the API returns after `clean_names()` — no
field translation is applied.

### Students

Key renames (v2 column → v3 column):

| v2 | v3 |
|---|---|
| `forename` | `first_name` |
| `middlename` | `middle_name` |
| `surname` | `surname` *(unchanged)* |
| `preferredname` | `preferred_name` |
| `previousname` | `previous_name` |
| `dob` | `date_of_birth` |
| `mobilephone` | `phone_mobile` |
| `secondaryphone` | `phone_home` |
| `email1` | `email_primary` |
| `email2` | `email_secondary` |
| `countryofbirth` | `country_of_birth` |
| `third_party_id` | `other_id_1` |
| `third_party_id2` | `other_id_2` |
| `disabilitydetails` | `disability_details` |
| `medicalcondition` | `medical_condition` |
| `bank_acc` | `bank_account` |

Fields present in v2 but **removed in v3**: `nsn`, `contacts`, `initials`,
`identitychecked`, `international`, `mainagent`, `austresidency`, `residency`,
`residentialstatus`, `passportnumber`, `passportexpiry`, `visatype`,
`visanumber`, `visaexpiry`, `ethnicity1`–`3`, `iwi1`–`3`, `lms_isuser`,
`validate_data_for`.

Fields **new in v3**: `pronoun`, `contact_id`, `user_name`,
`email_school`, `phone_work`, `homestay`, `primary_learning_style`,
`secondary_learning_style`, `student_identifier`,
`new_zealand_student_extension`, `australia_student_extension`.

### Enrolments

Key renames:

| v2 | v3 |
|---|---|
| `student_id` | `student` *(IRI → character)* |
| `intake_id` | `intake` *(IRI → character)* |
| `enrstartdate` | `start_date` |
| `enrenddate` | `end_date` |
| `enrstatus` | `enrolment_status` *(IRI → character)* |
| `enrstatusdate` | `enrolment_status_date` |
| `fundingsource` | `enrolment_payer_type_id` |
| `enrwithdrawalreason` | `withdrawal_reason` |
| `enrwithdrawaldate` | `withdrawal_date` |
| `enrcompletiondate` | `finished_date` |

### Enrolment components

| v2 | v3 |
|---|---|
| `compenrid` | `id` |
| `enrolid` | `enrolment` *(IRI → character)* |
| `compid` | `component` *(IRI → character)* |
| `compenrstartdate` | `start_date` |
| `compenrenddate` | `end_date` |
| `compenrduedate` | `due_date` |
| `compenrstatus` | `enrolment_status` *(IRI → character)* |
| `compenrwithdrawaldate` | `withdrawal_date` |
| `compenrcompletiondate` | `completion_date` |
| `compenrextensiondate` | `extension_date` |
| `compenrgrade` | *(via `enrolment_component_grades`)* |

Many v2 user-defined fields (`userfieldchar1` etc.) are replaced by
`custom_field_values` in v3.

### Intakes

| v2 | v3 |
|---|---|
| `intakeid` | `id` |
| `intakecode` | `code` |
| `intake_name` | `name` |
| `intakestartdate` | `start_date` |
| `intakeenddate` | `end_date` |
| `intakestatus` | `intake_status` *(nested object)* |
| `progid` | `programme` *(IRI → character)* |

### Programmes

| v2 | v3 |
|---|---|
| `progid` | `id` |
| `progcode` | `code` |
| `progtitle` | `title` |
| `progversion` | `version` |
| `progdescription` | `description` |
| `progefts` | *(removed; see duration fields)* |

---

## Updating code that references column names

Because column names differ, downstream code referencing specific column names
will need updating when switching from v2 to v3:

```{r}
# v2
students |> dplyr::select(id, forename, surname, dob, email1)

# v3
students |> dplyr::select(id, first_name, surname, date_of_birth, email_primary)
```

To support both versions in the same codebase, check `con$api_version`:

```{r}
name_col <- if (con$api_version == "v3") "first_name" else "forename"
students[[name_col]]
```

---

## Schema validation warnings

When the live API returns fewer fields than the package's built-in schema
snapshot (derived from the OpenAPI specs), selmaR emits a `cli_warn()` listing
the missing fields:

```
Warning: SELMA students (v2): 3 expected fields missing from response.
ℹ forename, dob, email1
ℹ The SELMA API may have changed. Check for a package update.
```

This is informational — the function still returns whatever data the API
provided. Suppress with `suppressWarnings()` once you have confirmed the
field removals are expected for your instance.
