---
title: "Getting Started with selmaR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with selmaR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

selmaR is an R client for the [SELMA](https://selma.co.nz/) student management
system API. SELMA is used by Private Training Establishments (PTEs) and Tertiary
Education Organisations (TEOs) across New Zealand.

This package handles:

- **Authentication** — connect once and all functions use your session
- **Pagination** — automatic Hydra JSON-LD page traversal
- **Data cleaning** — removes API artifacts, applies `clean_names()`, converts IDs to character
- **Caching** — optional RDS caching with configurable TTL

## Setup

### config.yml (Recommended)

Create a `config.yml` in your project root (**add it to `.gitignore`**):

```yaml
default:
  selma:
    base_url: "https://myorg.selma.co.nz/"
    email: "api@selma.co.nz"
    password: "your_password"
```

Then connect:

```{r}
library(selmaR)

selma_connect()
```

That's it. Every `selma_*()` function now uses your connection automatically.

### Environment Variables

Alternatively, set credentials in `.Renviron`:

```
SELMA_BASE_URL=https://myorg.selma.co.nz/
SELMA_EMAIL=api@selma.co.nz
SELMA_PASSWORD=your_password
```

### Direct Credentials

```{r}
selma_connect(
  base_url = "https://myorg.selma.co.nz/",
  email = "api@selma.co.nz",
  password = "your_password"
)
```

Credentials are resolved in priority order: direct arguments > config.yml > environment variables.

## Fetching Data

All fetch functions return tibbles with `clean_names()` applied:

```{r}
students   <- selma_students()
enrolments <- selma_enrolments()
intakes    <- selma_intakes()
components <- selma_components()
programmes <- selma_programmes()
```

### Filtering at the API Level

Most fetch functions accept optional filter parameters so you only download
what you need. Filtering happens server-side, reducing both API traffic and
memory usage:

```{r}
# Students by surname or forename
smiths <- selma_students(surname = "Smith")

# Components for a specific student or enrolment
comps <- selma_components(student_id = "123")
comps <- selma_components(enrol_id = "456")

# Intakes by programme, status, or date range
upcoming <- selma_intakes(prog_id = "42", start_after = "2025-01-01")

# Active programmes only
active_progs <- selma_programmes(status = "Active")

# Addresses for a specific student, contact, or organisation
addrs <- selma_addresses(student_id = "123")

# Contacts filtered by relationship
emergency <- selma_student_contacts(student_id = "123")

# Organisations by name or registration number
orgs <- selma_organisations(name = "Example Ltd")
```

See each function's help page (`?selma_students`, `?selma_intakes`, etc.)
for the full list of available filter parameters.

**Note:** When filter parameters are supplied, caching is automatically
bypassed — filtered results represent a subset and should not overwrite the
full cached dataset.

### Fetching a Single Record

Use `selma_get_one()` to retrieve a single record by ID from any endpoint:

```{r}
student <- selma_get_one("students", "123")
intake  <- selma_get_one("intakes", "456")
```

### Beyond the Core Entities

selmaR wraps the full SELMA API — notes, contacts, addresses, classes,
organisations, campuses, awards, custom fields, and all reference/lookup
tables:

```{r}
notes      <- selma_notes()
addresses  <- selma_addresses()
classes    <- selma_classes()
campuses   <- selma_campuses()
orgs       <- selma_organisations()
attempts   <- selma_component_attempts()
contacts   <- selma_contacts()
```

### Custom Fields

Retrieve custom field definitions and values:

```{r}
# Custom field tab definitions
fields <- selma_custom_fields()

# Custom field values for a specific student or enrolment
student_cf <- selma_student_custom_fields("123")
enrol_cf   <- selma_enrolment_custom_fields("456")
comp_cf    <- selma_component_custom_fields("456")
```

### Lookup / Reference Tables

SELMA exposes many reference code tables. selmaR wraps them all:

```{r}
ethnicities <- selma_ethnicities()
countries   <- selma_countries()
genders     <- selma_genders()
visa_types  <- selma_visa_types()
iwis        <- selma_nz_iwis()
# ... and many more — see the package reference index
```

### Intake Enrolments (Nested Data)

The intake enrolments endpoint returns nested JSON for a specific intake:

```{r}
ie <- selma_intake_enrolments(intake_id = 123)
```

## Joining Entities

Use the convenience join functions — no need to remember join keys:

```{r}
library(dplyr)

# Build a full student pipeline in one line
pipeline <- selma_student_pipeline(enrolments, students, intakes)

# Or use individual join helpers
selma_join_students(enrolments, students)     # enrolments + student details
selma_join_intakes(enrolments, intakes)       # enrolments + intake dates
selma_join_components(components, enrolments) # components + enrolment info
selma_join_programmes(intakes, programmes)    # intakes + programme names
selma_join_notes(notes, students)             # notes + student details
selma_join_addresses(addresses, students)     # addresses + student details
selma_join_classes(classes, campuses)         # classes + campus locations
selma_join_attempts(attempts, components)     # attempts + component details

# Full component pipeline: components + enrolments + students + intakes
full <- selma_component_pipeline(components, enrolments, students, intakes)
```

## Using the Cache

Enable caching to avoid slow API calls on repeated runs:
```{r}
students <- selma_students(cache = TRUE, cache_dir = "data", cache_hours = 24)
```

## URL Builders

Generate links to SELMA records:

```{r}
selma_student_url(123, "https://myorg.selma.co.nz/")
#> "https://myorg.selma.co.nz/en/admin/student/123/1"

selma_enrolment_url(456, "https://myorg.selma.co.nz/")
#> "https://myorg.selma.co.nz/en/admin/enrolment/456/1"
```

## Status Code Constants

Use the exported constants for filtering:

```{r}
# Active funded enrolments
enrolments |>
  filter(enrstatus %in% c(SELMA_STATUS_CONFIRMED, SELMA_STATUS_COMPLETED))

# All funded statuses (including withdrawn)
SELMA_ALL_FUNDED_STATUSES
#> [1] "C"  "FC" "FI" "WR" "WS"
```

## Generic Endpoint Access

For any collection endpoint, you can use `selma_get()` directly:

```{r}
data <- selma_get(endpoint = "some_endpoint")
```

For a single record by ID, use `selma_get_one()`:

```{r}
record <- selma_get_one("students", "123")
```
