---
title: "Phone Number Normalisation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Phone Number Normalisation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

SELMA stores phone numbers in various formats. `standardize_phone()` uses the
[dialvalidator](https://cran.r-project.org/package=dialvalidator) package
(powered by Google's libphonenumber metadata) to parse, validate, and format
phone numbers to E.164 international format.

By default, numbers without a country code are tried as NZ first, then AU.
This handles the common case of SELMA instances with students across both
countries.

## Examples

### New Zealand (+64)

```{r}
library(selmaR)

standardize_phone("021 123 4567")     # NZ mobile with leading 0
#> "+64211234567"

standardize_phone("0211234567")       # No spaces
#> "+64211234567"

standardize_phone("+64211234567")     # Already international
#> "+64211234567"

standardize_phone("211234567")        # Mobile without leading 0
#> "+64211234567"
```

### Australia (+61)

AU numbers with a `+61` prefix work directly. Local AU numbers (e.g. `04xx`)
are automatically detected via the NZ-then-AU fallback — `04` is not a valid
NZ prefix, so dialvalidator correctly parses it as Australian:

```{r}
standardize_phone("+61412345678")     # Already international
#> "+61412345678"

standardize_phone("0412345678")       # AU local — detected automatically
#> "+61412345678"
```

### Invalid Numbers

Numbers that fail validation in all regions are returned as `NA`:

```{r}
standardize_phone("1234")            # Too short
#> NA

standardize_phone(NA)                # NA passthrough
#> NA

standardize_phone("")                # Empty string
#> NA
```

## Vectorised Usage

Use `standardize_phones()` for vectors — handles mixed NZ and AU numbers:

```{r}
phones <- c("021 123 4567", "0412345678", "+61412345678", NA)
standardize_phones(phones)
#> [1] "+64211234567" "+61412345678" "+61412345678" NA
```

## Use with dplyr

```{r, eval = FALSE}
library(dplyr)

students |>
  mutate(
    mobile_std = standardize_phone(mobilephone),
    secondary_std = standardize_phone(secondaryphone)
  )
```

## Custom Regions

Override the default fallback regions if your student population differs:

```{r, eval = FALSE}
# Only try AU (no NZ fallback)
standardize_phones(phones, default_regions = "AU")

# Try NZ, AU, then UK
standardize_phones(phones, default_regions = c("NZ", "AU", "GB"))
```
